### 2020-12-13 written/eLifeRevision/Bittner_eLife2020.tex
EPI as VI with mf gauss llhd

%\begin{equation}\label{eq:EP}
 %\mathbb{E}\left[T(x) \right] ~~\triangleq~~ \mathbb{E}\begin{bmatrix} \omega_{\text{f1}}(x) \\ \vdots \\ (\omega_{\text{f1}}(x) - 0.53)^2 \\ \vdots \end{bmatrix} ~~=~~  
 %\begin{bmatrix} 0.53 \\ \vdots \\ 0.025^2 \\ \vdots \end{bmatrix} ~~\triangleq~~ \mu,
 %\end{equation}
\begin{equation}\label{eq:EP}
\begin{split}
\mathcal{X} ~~:~~ \mathbb{E}_{\mathbf{z}}\left[T(\mathbf{x}; \mathbf{z})\right] &~~\triangleq~~  \mathbb{E}_{\mathbf{z}}\begin{bmatrix} \omega_{\text{f1}}(\mathbf{x}; \mathbf{z}) \\ \omega_{\text{f2}}(\mathbf{x}; \mathbf{z}) \\ \vdots \end{bmatrix}  ~~=~~  \begin{bmatrix} 0.53 \\ 0.53 \\ \vdots \end{bmatrix} ~~\triangleq~~ \bm{\mu}  \\ 
 \text{Var}_{\mathbf{z}}\left[T(\mathbf{x}; \mathbf{z})\right] &~~\triangleq~~ \text{Var}_{\mathbf{z}}\begin{bmatrix} \omega_{\text{f1}}(\mathbf{x}; \mathbf{z}) \\ \omega_{\text{f2}}(\mathbf{x}; \mathbf{z}) \\ \vdots \end{bmatrix}  ~~=~~  \begin{bmatrix} 0.025^2 \\ 0.025^2 \\ \vdots \end{bmatrix} ~~\triangleq~~ \bm{\sigma}^2 .
\end{split}
\end{equation}

@@

Third, we perform emergent property inference: we find a distribution over parameter configurations $\mathbf{z}$, and insist that samples from this distribution produce the emergent property; in other words, they obey the constraints introduced in Equation \ref{eq:EP}.  
This distribution will be chosen from a family of probability distributions $\mathcal{Q} = \left\{ q_{\bm{\theta}}(\mathbf{z}) : \bm{\theta} \in \Theta \right\}$, defined by a deep generative distribution of the normalizing flow class \cite{rezende2015variational, dinh2016density, papamakarios2017masked} -- neural networks which transform a simple distribution into a suitably complicated distribution (as is needed here).  
This deep distribution is represented in Figure \ref{fig:STG}B (see Section \ref{methods_EPI}).  
Then, mathematically, we must solve the following optimization program: 
%\begin{equation} \label{eq:EPI}
%\begin{split}
%&\argmax_{q_{\bm{\theta}} \in \mathcal{Q}} H(q_{\bm{\theta}}(\mathbf{z})) \\
%&\text{  s.t.  } \mathbb{E}_{\mathbf{z} \sim q_{\bm{\theta}}}\left[ \mathbb{E}_{\mathbf{x} \sim p(\mathbf{x} \mid \mathbf{z})}\left[T(\mathbf{x})\right] \right] = \bm{\mu}, \\
%\end{split}
%\end{equation}
\begin{equation} \label{eq:EPI}
\begin{split}
q_{\bm{\theta}}(\mathbf{z} \mid \mathcal{X}) ~~=~~ &\argmax_{q_{\bm{\theta}} \in \mathcal{Q}} H(q_{\bm{\theta}}(\mathbf{z})) \\
&\text{  s.t.  } ~\mathcal{X}:~~ \mathbb{E}_{\mathbf{z}}\left[T(\mathbf{x}; \mathbf{z})\right] = \bm{\mu}, \text{Var}_{\mathbf{z}}\left[T(\mathbf{x}; \mathbf{z})\right] = \bm{\sigma}^2 \\
\end{split}
\end{equation}
%where $T(\mathbf{x}), \bm{\mu}$ are defined as in Equation \ref{eq:EP}, and $p(\mathbf{x} | \mathbf{z})$ is the intractable distribution of data from the model, $\mathbf{x}$, given that model's parameters $\mathbf{z}$ (we access samples from this distribution by running the model forward).
where $T(\mathbf{x}, \mathbf{z}), \bm{\mu}$, and $\bm{\sigma}$ are defined as in Equation \ref{eq:EP}.
%The purpose of each element in this program is detailed in Figure \ref{fig:STG}D.
Finally, we recognize that many distributions in $\mathcal{Q}$ will respect the emergent property constraints, so we select that which has maximum entropy.
This principle, captured in Equation \ref{eq:EPI} by the primal objective $H$, identifies parameter distributions with minimal assumptions beyond some chosen structure \cite{jaynes1957information, elsayed2017structure, loaiza2017maximum, savin2017maximum}.
Such a normative principle of maximum entropy, which is also that of Bayesian inference, naturally fits with our scientific objective of reasoning about parametric sensitivity and robustness. 
The recovered distribution of EPI is as variable as possible along each parametric manifold such that it produces the emergent property.

@@

Using EPI, we inferred the distribution of $dh$ shown in Figure \ref{fig:V1_EPI}C producing $\mathcal{B}(\alpha, y)$.  
Columns correspond to inferred distributions of excitatory ($\alpha=E$, red), parvalbumin ($\alpha=P$, blue), somatostatin ($\alpha=S$, red) and VIP ($\alpha=V$, green) neuron-type response increases, while each row corresponds to increase amounts of $y \in \{0.1, 0.5\}$.  For each pair of parameters, we show the two-dimensional marginal distribution of samples colored by $\log q_\theta(dh \mid \mathcal{B}(\alpha, y))$.  The inferred distributions immediately suggest four hypotheses: \\

@@

\begin{figure}
\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=4cm}}]{figure}[\FBwidth]
{\caption{Confirming EPI generated hypotheses in V1. 
A. Differential responses $\delta x_E$ by the E-population to changes in individual input $\delta h_\alpha \hat{u}_\alpha$ away from the mode of the EPI distribution $dh^*$. 
B-D Same plots for the P-, S-, and V-populations.  
Labels H1, H2, H3, and H4 indicate which curves confirm which hypotheses.}\label{fig:V1_HT}}
{\includegraphics[scale=0.6]{figures/fig3/fig3.pdf}}
\end{figure}
We evaluate these hypotheses by taking perturbations in individual neuron-type input $\delta h_\alpha$ away from the modes of the inferred distributions at $y=0.1$
\begin{equation}
dh^* = z^* = \argmax_{z} \log q_\theta(z \mid \mathcal{B}(\alpha, 0.1)).
\end{equation}
Here $\delta x_{\alpha}$ is the change in steady state response of the system with input $h = b + dh^* + \delta h_\alpha \hat{u}_\alpha$ compared to $h = b + dh^*$, where $\hat{u}_\alpha$ is a unit vector in the dimension of $\alpha$. 
The EPI-generated  hypotheses are confirmed (for details, see Section \ref{methods_V1}):

@@

\subsection{Emergent property inference (EPI)}\label{methods_EPI}
Emergent property inference (EPI) learns distributions of theoretical model parameters that produce emergent properties of interest by combining ideas from maximum entropy flow networks (MEFNs) \cite{loaiza2017maximum} and likelihood-free variational inference (LFVI) \cite{tran2017hierarchical}.  
Consider model parameterization $z$ and data $x$ which has an intractable likelihood $p(x \mid z)$ defined by a model simulator of which samples are available $x \sim p(x \mid z)$.  
EPI optimizes a distribution $q_\theta(z)$ (itself parameterized by $\theta$) of model parameters $z$ to produce an emergent property of interest $\mathcal{B}$,
\begin{equation}
\mathcal{B} ~~\triangleq~~  \mathbb{E}_{z \sim q_\theta}\left[ \mathbb{E}_{x\sim p(x \mid z)}\left[T(x)\right] \right] = \mu.
\end{equation}
Precisely, the emergent property statistics $T(x)$ must equal the emergent property values $\mu$, in expectation over the EPI distribution of parameters $q_\theta(z)$ and the distribution of simulated activity $p(x \mid z)$.  
This is a viable way to represent emergent properties in theoretical models, as we have demonstrated in the main text, and enables the EPI optimization.

With EPI, we use deep probability distributions to learn flexible approximations to model parameter distributions $q_\theta(z)$.
 In deep probability distributions, a simple random variable $w \sim q_0(w)$ is mapped deterministically via a sequence of deep neural network layers ($f_1$, .. $f_l$) parameterized by weights and biases $\theta$ to the support of the distribution of interest:
\begin{equation}
z = f_{\theta}(\omega) = f_l(..f_1(w)).
\end{equation}
Given a simulator defined by a theoretical model $x \sim p(x \mid z)$ and some emergent property of interest $\mathcal{B}$, $q_\theta(z)$ is optimized via the neural network parameters $\theta$ to find a maximally entropic distribution $q_{\theta}^*$ within the deep variational family $\mathcal{Q}$ producing the emergent property:
\begin{equation} \label{eq:opt}
\begin{split}
q_\theta^*(z) &= \argmax_{q_\theta \in Q} H(q_\theta(z)) \\
 &  \text{s.t.  } \mathbb{E}_{z \sim q_\theta}\left[ \mathbb{E}_{x\sim p(x \mid z)}\left[T(x)\right] \right] = \mu. \\
 \end{split}
\end{equation} 
Since we are optimizing parameters $\theta$ of our deep probability distribution with respect to the entropy $H(q_\theta(z))$, we must take gradients with respect to the log probability density of samples from the deep probability distribution.  
Entropy of $q_\theta(z)$ can be expressed as an expectation of the negative log density of parameter samples $z$ over the randomness in the parameterless initial distribution $q_0$:
\begin{equation}
H(q_\theta(z)) = \int - q_\theta(z) \log(q_\theta(z)) dz = \mathbb{E}_{z \sim q_\theta}\left[-\log(q_\theta(z)) \right] = \mathbb{E}_{w \sim q_0}\left[-\log(q_\theta(f_\theta(w))) \right].
\end{equation}
Thus, the gradient of the entropy of the deep probability distribution can be estimated as an average of gradients of the log density of samples $z$:
\begin{equation}
\nabla_\theta H(q_\theta(z)) = \mathbb{E}_{w \sim q_0}\left[- \nabla_\theta \log(q_\theta(f_\theta(w))) \right].
\end{equation}
In EPI, MEFNs are purposed towards variational learning of model parameter distributions.  
A closely related methodology, variational inference, uses optimization to approximate posterior distributions \cite{blei2017variational}.
Standard methods like stochastic gradient variational Bayes \cite{kingma2013auto} or black box variational inference \cite{ranganath2014black} simply do not work for inference in theoretical models of neural circuits, since they require tractable likelihoods $p(x \mid z)$.
Work on likelihood-free variational inference (LFVI) \cite{tran2017hierarchical}, which like EPI seeks to do inference in models with intractable likelihoods, employs an addtional deep neural network as a ratio estimator, enabling an estimation of the optimization objective for variational inference.
Like LFVI, EPI can be framed as variational inference (see Section \ref{methods_VI}).
But, unlike LFVI, EPI uses a single deep network to learn a distribution and is optimized to produce an emergent property, rather than condition on data points.  
Optimizing the EPI objective is a technological challenge, the details of which we elaborate in Section \ref{methods_AL_opt}.  
Before going through those details, we ground this optimization in a toy example.

 We note that, during our preparation and early presentation of this work \cite{bittner2019degenerate, bittner2019examining}, another work has arisen with broadly similar goals: bringing statistical inference to mechanistic models of neural circuits (\cite{nonnenmacher2018sbi, desitler2019statistical, gonccalves2019training}, preprint posted simultaneously with this preprint).
We are encouraged by this general problem being recognized by others in the community, and we emphasize that these works offer complementary neuroscientific contributions (different theoretical models of focus) and use different technical methodologies (ours is built on our prior work \cite{loaiza2017maximum}, theirs similarly \cite{LueckmannGoncalves_17}).
These distinct methodologies and scientific investigations emphasize the increased importance and timeliness of both works.

\subsubsection{Example: 2D LDS}\label{methods_2DLDS}
To gain intuition for EPI, consider a two-dimensional linear dynamical system (2D LDS) model (Fig. S1A):
\begin{equation} 
\tau \frac{dx}{dt} = Ax
\end{equation}
with
\begin{equation}
A = \begin{bmatrix} a_1 & a_2 \\ a_3 & a_4 \end{bmatrix}.
\end{equation}
To run EPI with the dynamics matrix elements as the free parameters $z = [ a_1, a_2, a_3, a_4]$ (fixing $\tau=1$), the emergent property statistics $T(x)$ were chosen to contain the first and second moments of the oscillatory frequency, $\frac{\text{imag}(\lambda_1)}{2 \pi}$, and the growth/decay factor, $\text{real}(\lambda_1)$, of the oscillating system. 
 $\lambda_1$ is the eigenvalue of greatest real part when the imaginary component is zero, and alternatively of positive imaginary component when the eigenvalues are complex conjugate pairs.  
To learn the distribution of real entries of $A$ that produce a band of oscillating systems around 1Hz, we formalized this emergent property as $\text{real}(\lambda_1)$ having mean zero with variance $0.25^2$, and the oscillation frequency $2 \pi \text{imag}(\lambda_1)$ having mean $\omega = 1$ Hz with variance (0.1Hz)$^2$:
\begin{equation}
 \mathbb{E}\left[T(x) \right] ~~ \triangleq ~~ \mathbb{E} \begin{bmatrix} \text{real}(\lambda_1) \\ \text{imag}(\lambda_1) \\ (\text{real}(\lambda_1)-0)^2  \\ (\text{imag}(\lambda_1)-2 \pi \omega)^2 \end{bmatrix} = \begin{bmatrix} 0.0 \\ 2 \pi \omega \\ 0.25^2 \\ (2 \pi 0.1)^2 \end{bmatrix} ~~ \triangleq ~~ \mu.
 \end{equation} 
\begin{figure}
\begin{center}
\includegraphics[scale=0.58]{figures/figS1/figS1.pdf}
\end{center}
\begin{flushleft}
{\small Fig. S1: A. Two-dimensional linear dynamical system model, where real entries of the dynamics matrix $A$ are the parameters.  
B. The EPI distribution for a two-dimensional linear dynamical system with $\tau=1$ that produces an average of 1Hz oscillations with some small amount of variance.  Dashed lines indicate the parameter axes. 
C. Entropy throughout the optimization.  
At the beginning of each augmented Lagrangian epoch (2,000 iterations), the entropy dipped due to the shifted optimization manifold where emergent property constraint satisfaction is increasingly weighted.  
D. Emergent property moments throughout optimization.  
At the beginning of each augmented Lagrangian epoch, the emergent property moments adjust closer to their constraints.}
\end{flushleft}
\end{figure}

Unlike the models we presented in the main text, this model admits an analytical form for the mean emergent property statistics given parameter $z$, since the eigenvalues can be calculated using the quadratic formula: 
\begin{equation}
\lambda = \frac{(\frac{a_1 + a_4}{\tau}) \pm \sqrt{(\frac{a_1+a_4}{\tau})^2 + 4(\frac{a_2 a_3 - a_1 a_4}{\tau})}}{2}.
\end{equation}

Importantly, even though $\mathbb{E}_{x\sim p(x \mid z)}\left[T(x)\right]$ is calculable directly via a closed form function and does not require simulation, we cannot derive the distribution $q^*_\theta$ directly.  
This fact is due to the formally hard problem of the backward mapping: finding the natural parameters $\eta$ from the mean parameters $\mu$ of an exponential family distribution \cite{wainwright2008graphical}.  
Instead, we used EPI to approximate this distribution (Fig. S1B). We used a real-NVP normalizing flow architecture with four masks, two neural network layers of 15 units per mask, with batch normalization momentum 0.99, mapped onto a support of $z_i \in \left[-10, 10 \right]$. (see Section \ref{methods_NF}).

>>>>

%\begin{equation}\label{eq:EP}
 %\mathbb{E}\left[T(x) \right] ~~\triangleq~~ \mathbb{E}\begin{bmatrix} \omega_{\text{f1}}(x) \\ \vdots \\ (\omega_{\text{f1}}(x) - 0.53)^2 \\ \vdots \end{bmatrix} ~~=~~  
 %\begin{bmatrix} 0.53 \\ \vdots \\ 0.025^2 \\ \vdots \end{bmatrix} ~~\triangleq~~ \mu,
 %\end{equation}
\begin{equation}\label{eq:EP}
\begin{split}
\mathcal{X} ~~:~~ \mathbb{E}_{\mathbf{z},\mathbf{x}}\left[T(\mathbf{x}; \mathbf{z})\right] &~~\triangleq~~  \mathbb{E}_{\mathbf{z},\mathbf{x}}\begin{bmatrix} \omega_{\text{f1}}(\mathbf{x}; \mathbf{z}) \\ \omega_{\text{f2}}(\mathbf{x}; \mathbf{z}) \\ \vdots \end{bmatrix}  ~~=~~  \begin{bmatrix} 0.53 \\ 0.53 \\ \vdots \end{bmatrix} ~~\triangleq~~ \bm{\mu}  \\ 
 \text{Var}_{\mathbf{z},\mathbf{x}}\left[T(\mathbf{x}; \mathbf{z})\right] &~~\triangleq~~ \text{Var}_{\mathbf{z},\mathbf{x}}\begin{bmatrix} \omega_{\text{f1}}(\mathbf{x}; \mathbf{z}) \\ \omega_{\text{f2}}(\mathbf{x}; \mathbf{z}) \\ \vdots \end{bmatrix}  ~~=~~  \begin{bmatrix} 0.025^2 \\ 0.025^2 \\ \vdots \end{bmatrix} ~~\triangleq~~ \bm{\sigma}^2 .
\end{split}
\end{equation}

@@

Third, we perform emergent property inference: we find a distribution over parameter configurations $\mathbf{z}$, and insist that samples from this distribution produce the emergent property; in other words, they obey the constraints introduced in Equation \ref{eq:EP}.  
This distribution will be chosen from a family of probability distributions $\mathcal{Q} = \left\{ q_{\bm{\theta}}(\mathbf{z}) : \bm{\theta} \in \Theta \right\}$, defined by a deep generative distribution of the normalizing flow class \cite{rezende2015variational, dinh2016density, papamakarios2017masked} -- neural networks which transform a simple distribution into a suitably complicated distribution (as is needed here).  
This deep distribution is represented in Figure \ref{fig:STG}B (see Section \ref{methods_EPI}).  
Then, mathematically, we must solve the following optimization program: 
%\begin{equation} \label{eq:EPI}
%\begin{split}
%&\argmax_{q_{\bm{\theta}} \in \mathcal{Q}} H(q_{\bm{\theta}}(\mathbf{z})) \\
%&\text{  s.t.  } \mathbb{E}_{\mathbf{z} \sim q_{\bm{\theta}}}\left[ \mathbb{E}_{\mathbf{x} \sim p(\mathbf{x} \mid \mathbf{z})}\left[T(\mathbf{x})\right] \right] = \bm{\mu}, \\
%\end{split}
%\end{equation}
\begin{equation} \label{eq:EPI}
\begin{split}
q_{\bm{\theta}}(\mathbf{z} \mid \mathcal{X}) ~~=~~ &\argmax_{q_{\bm{\theta}} \in \mathcal{Q}} H(q_{\bm{\theta}}(\mathbf{z})) \\
&\text{  s.t.  } ~\mathcal{X}:~~ \mathbb{E}_{\mathbf{z},\mathbf{x}}\left[T(\mathbf{x}; \mathbf{z})\right] = \bm{\mu}, \text{Var}_{\mathbf{z},\mathbf{x}}\left[T(\mathbf{x}; \mathbf{z})\right] = \bm{\sigma}^2 \\
\end{split}
\end{equation}
%where $T(\mathbf{x}), \bm{\mu}$ are defined as in Equation \ref{eq:EP}, and $p(\mathbf{x} | \mathbf{z})$ is the intractable distribution of data from the model, $\mathbf{x}$, given that model's parameters $\mathbf{z}$ (we access samples from this distribution by running the model forward).
where $T(\mathbf{x}, \mathbf{z}), \bm{\mu}$, and $\bm{\sigma}$ are defined as in Equation \ref{eq:EP}.
%The purpose of each element in this program is detailed in Figure \ref{fig:STG}D.
Finally, we recognize that many distributions in $\mathcal{Q}$ will respect the emergent property constraints, so we select that which has maximum entropy.
This principle, captured in Equation \ref{eq:EPI} by the primal objective $H$, identifies parameter distributions with minimal assumptions beyond some chosen structure \cite{jaynes1957information, elsayed2017structure, loaiza2017maximum, savin2017maximum}.
Such a normative principle of maximum entropy, which is also that of Bayesian inference, naturally fits with our scientific objective of reasoning about parametric sensitivity and robustness. 
The recovered distribution of EPI is as variable as possible along each parametric manifold such that it produces the emergent property.

@@

Using EPI, we inferred the distribution of $dh$ shown in Figure \ref{fig:V1_EPI}C producing $\mathcal{B}(\alpha, y)$.  
Columns correspond to inferred distributions of excitatory ($\alpha=E$, red), parvalbumin ($\alpha=P$, blue), somatostatin ($\alpha=S$, red) and VIP ($\alpha=V$, green) neuron-type response increases, while each row corresponds to increase amounts of $y \in \{0.1, 0.5\}$.  For each pair of parameters, we show the two-dimensional marginal distribution of samples colored by $\log q_{\bm{\theta}}(dh \mid \mathcal{B}(\alpha, y))$.  The inferred distributions immediately suggest four hypotheses: \\

@@

\begin{figure}
\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=4cm}}]{figure}[\FBwidth]
{\caption{Confirming EPI generated hypotheses in V1. 
A. Differential responses $\delta x_E$ by the E-population to changes in individual input $\delta h_\alpha \hat{u}_\alpha$ away from the mode of the EPI distribution $dh^*$. 
B-D Same plots for the P-, S-, and V-populations.  
Labels H1, H2, H3, and H4 indicate which curves confirm which hypotheses.}\label{fig:V1_HT}}
{\includegraphics[scale=0.6]{figures/fig3/fig3.pdf}}
\end{figure}
We evaluate these hypotheses by taking perturbations in individual neuron-type input $\delta h_\alpha$ away from the modes of the inferred distributions at $y=0.1$
\begin{equation}
dh^* = z^* = \argmax_{z} \log q_{\bm{\theta}}(z \mid \mathcal{B}(\alpha, 0.1)).
\end{equation}
Here $\delta x_{\alpha}$ is the change in steady state response of the system with input $h = b + dh^* + \delta h_\alpha \hat{u}_\alpha$ compared to $h = b + dh^*$, where $\hat{u}_\alpha$ is a unit vector in the dimension of $\alpha$. 
The EPI-generated  hypotheses are confirmed (for details, see Section \ref{methods_V1}):

@@

\subsection{Emergent property inference (EPI)}\label{methods_EPI}
%Emergent property inference (EPI) learns distributions of theoretical model parameters that produce emergent properties of interest by combining ideas from maximum entropy flow networks (MEFNs) \cite{loaiza2017maximum} and likelihood-free variational inference (LFVI) \cite{tran2017hierarchical}.  
%Consider model parameterization $\mathbf{z}$ and data $\mathbf{x}$ which has an intractable likelihood $p(\mathbf{x} \mid \mathbf{z})$ defined by a model simulator of which samples are available $\mathbf{x} \sim p(\mathbf{x} \mid \mathbf{z})$.  
%EPI optimizes a distribution $q_{\bm{\theta}}(\mathbf{z})$ (itself parameterized by $\bm{\theta}$) of model parameters $\mathbf{z}$ to produce an emergent property of interest $\mathcal{B}$,
%\begin{equation}
%\mathcal{B} ~~\triangleq~~  \mathbb{E}_{\mathbf{z} \sim q_{\bm{\theta}}}\left[ \mathbb{E}_{\mathbf{x}\sim p(\mathbf{x} \mid \mathbf{z})}\left[T(\mathbf{x})\right] \right] = \mu.
%\end{equation}
%Precisely, the emergent property statistics $T(\mathbf{x})$ must equal the emergent property values $\bm{\mu}$, in expectation over the EPI distribution of parameters $q_{\bm{\theta}}(\mathbf{z})$ and the distribution of simulated activity $p(\mathbf{x} \mid \mathbf{z})$.  
%This is a viable way to represent emergent properties in theoretical models, as we have demonstrated in the main text, and enables the EPI optimization.Emergent property inference (EPI) learns distributions of theoretical model parameters that produce emergent properties of interest by combining ideas from maximum entropy flow networks (MEFNs) \cite{loaiza2017maximum} and likelihood-free variational inference (LFVI) \cite{tran2017hierarchical}.  
Consider model parameterization $\mathbf{z}$ and data $\mathbf{x}$ which has an intractable likelihood $p(\mathbf{x} \mid \mathbf{z})$ defined by a model simulator of which samples are available $\mathbf{x} \sim p(\mathbf{x} \mid \mathbf{z})$.  
EPI optimizes a distribution $q_{\bm{\theta}}(\mathbf{z})$ (itself parameterized by $\bm{\theta}$) of model parameters $\mathbf{z}$ to produce an emergent property of interest $\mathcal{X}$ defined by the means and variances of emergent property statistics $f(\mathbf{x}; \mathbf{z})$
 \begin{equation}
\mathcal{X}: \mathbb{E}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\mu}, \text{Var}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\sigma}^2.
\end{equation}
Precisely, the emergent property statistics $f(\mathbf{x})$ must have means $\bm{\mu}$ and variances $\bm{\sigma}^2$  over the EPI distribution of parameters $q_{\bm{\theta}}(\mathbf{z})$ and stochasticity of the data given the parameters defined by the model $p(\mathbf{x} \mid \mathbf{z})$.  
This is a viable way to represent emergent properties in theoretical models, as we have demonstrated in the main text, and enables the EPI optimization.

%With EPI, we use deep probability distributions to learn flexible approximations to model parameter distributions $q_{\bm{\theta}}(\mathbf{z})$.
% In deep probability distributions, a simple random variable $\mathbf{z}_0 \sim q_0(\mathbf{z}_0)$ is mapped deterministically via a sequence of deep neural network layers ($g_1$, .. $g_l$) parameterized by weights and biases $\bm{\theta}$ to the support of the distribution of interest:
%\begin{equation}
%\mathbf{z} = g_{\bm{\theta}}(\mathbf{z}_0) = g_l(..g_1(\mathbf{z}_0)) \sim q_{\bm{\theta}}(\mathbf{z}).
%\end{equation}
%Given a simulator defined by a theoretical model $\mathbf{x} \sim p(\mathbf{x} \mid \mathbf{z})$ and some emergent property of interest $\mathcal{X}$, $q_{\bm{\theta}}(\mathbf{z})$ is optimized via the neural network parameters $\bm{\theta}$ to find a maximally entropic distribution $q_{\bm{\theta}}^*$ within the deep variational family $\mathcal{Q}$ producing the emergent property:
%\begin{equation} \label{eq:opt}
%\begin{split}
%q_{\bm{\theta}}^*(\mathbf{z}) &= \argmax_{q_{\bm{\theta}} \in Q} H(q_{\bm{\theta}}(\mathbf{z})) \\
% &  \text{s.t.  } \mathbb{E}_{\mathbf{z},\mathbf{x}}\left[T(\mathbf{x}; \mathbf{z})\right] = \bm{\mu}, \text{Var}_{\mathbf{z},\mathbf{x}}\left[T(\mathbf{x}; \mathbf{z})\right] = \bm{\sigma}^2. \\
% \end{split}
%\end{equation} 
%Since we are optimizing parameters $\bm{\theta}$ of our deep probability distribution with respect to the entropy $H(q_{\bm{\theta}}(\mathbf{z}))$, we must take gradients with respect to the log probability density of samples from the deep probability distribution.  
%Entropy of $q_{\bm{\theta}}(\mathbf{z})$ can be expressed as an expectation of the negative log density of parameter samples $\mathbf{z}$ over the randomness in the parameterless initial distribution $q_0$:
%\begin{equation}
%H(q_{\bm{\theta}}(\mathbf{z})) = \int - q_{\bm{\theta}}(\mathbf{z}) \log(q_{\bm{\theta}}(\mathbf{z})) d\mathbf{z} = \mathbb{E}_{\mathbf{z} \sim q_{\bm{\theta}}}\left[-\log(q_{\bm{\theta}}(\mathbf{z})) \right] = \mathbb{E}_{w \sim q_0}\left[-\log(q_{\bm{\theta}}(f_{\bm{\theta}}(w))) \right].
%\end{equation}
%Thus, the gradient of the entropy of the deep probability distribution can be estimated as an average of gradients of the log density of samples $\mathbf{z}$:
%\begin{equation}
%\nabla_{\bm{\theta}} H(q_{\bm{\theta}}(\mathbf{z})) = \mathbb{E}_{w \sim q_0}\left[- \nabla_{\bm{\theta}} \log(q_{\bm{\theta}}(f_{\bm{\theta}}(w))) \right].
%\end{equation}
%In EPI, MEFNs are purposed towards variational learning of model parameter distributions.  
%A closely related methodology, variational inference, uses optimization to approximate posterior distributions \cite{blei2017variational}.
%Standard methods like stochastic gradient variational Bayes \cite{kingma2013auto} or black box variational inference \cite{ranganath2014black} simply do not work for inference in theoretical models of neural circuits, since they require tractable likelihoods $p(\mathbf{x} \mid \mathbf{z})$.
%Work on likelihood-free variational inference (LFVI) \cite{tran2017hierarchical}, which like EPI seeks to do inference in models with intractable likelihoods, employs an addtional deep neural network as a ratio estimator, enabling an estimation of the optimization objective for variational inference.
%Like LFVI, EPI can be framed as variational inference (see Section \ref{methods_VI}).
%But, unlike LFVI, EPI uses a single deep network to learn a distribution and is optimized to produce an emergent property, rather than condition on data points.  
%Optimizing the EPI objective is a technological challenge, the details of which we elaborate in Section \ref{methods_AL_opt}.  
%Before going through those details, we ground this optimization in a toy example.

With EPI, we use deep probability distributions to learn flexible approximations to model parameter distributions $q_{\bm{\theta}}(\mathbf{z})$.
 In deep probability distributions, a simple random variable $\mathbf{z}_0 \sim q_0(\mathbf{z}_0)$ is mapped deterministically via a sequence of deep neural network layers ($g_1$, .. $g_l$) parameterized by weights and biases $\bm{\theta}$ to the support of the distribution of interest:
\begin{equation}
\mathbf{z} = g_{\bm{\theta}}(\mathbf{z}_0) = g_l(..g_1(\mathbf{z}_0)) \sim q_{\bm{\theta}}(\mathbf{z}).
\end{equation}
Given a simulator defined by a theoretical model $\mathbf{x} \sim p(\mathbf{x} \mid \mathbf{z})$ and some emergent property of interest $\mathcal{X}$, $q_{\bm{\theta}}(\mathbf{z})$ is optimized via the neural network parameters $\bm{\theta}$ to find a maximally entropic distribution $q_{\bm{\theta}}^*$ within the deep variational family $\mathcal{Q}$ producing the emergent property:
\begin{equation} \label{eq:opt}
\begin{split}
q_{\bm{\theta}}^*(\mathbf{z}) &= \argmax_{q_{\bm{\theta}} \in Q} H(q_{\bm{\theta}}(\mathbf{z})) \\
 &  \text{s.t.  } \mathbb{E}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\mu}, \text{Var}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\sigma}^2. \\
 \end{split}
\end{equation} 
Since we are optimizing parameters $\bm{\theta}$ of our deep probability distribution with respect to the entropy $H(q_{\bm{\theta}}(\mathbf{z}))$, we must take gradients with respect to the log probability density of samples from the deep probability distribution.  
Entropy of $q_{\bm{\theta}}(\mathbf{z})$ can be expressed as an expectation of the negative log density of parameter samples $\mathbf{z}$ over the randomness in the parameterless initial distribution $q_0$:
\begin{equation}
H(q_{\bm{\theta}}(\mathbf{z})) = \int - q_{\bm{\theta}}(\mathbf{z}) \log(q_{\bm{\theta}}(\mathbf{z})) d\mathbf{z} = \mathbb{E}_{\mathbf{z} \sim q_{\bm{\theta}}}\left[-\log(q_{\bm{\theta}}(\mathbf{z})) \right] = \mathbb{E}_{\mathbf{z}_0 \sim q_0}\left[-\log(q_{\bm{\theta}}(g_{\bm{\theta}}(\mathbf{z}_0))) \right].
\end{equation}
Thus, the gradient of the entropy of the deep probability distribution can be estimated as an average of gradients of the log density of samples $\mathbf{z}$:
\begin{equation}
\nabla_{\bm{\theta}} H(q_{\bm{\theta}}(\mathbf{z})) = \mathbb{E}_{\mathbf{z}_0 \sim q_0}\left[- \nabla_{\bm{\theta}} \log(q_{\bm{\theta}}(g_{\bm{\theta}}(\mathbf{z}_0))) \right].
\end{equation}
In EPI, MEFNs are purposed towards variational learning of model parameter distributions.

% We note that, during our preparation and early presentation of this work \cite{bittner2019degenerate, bittner2019examining}, another work has arisen with broadly similar goals: bringing statistical inference to mechanistic models of neural circuits (\cite{nonnenmacher2018sbi, desitler2019statistical, gonccalves2019training}, preprint posted simultaneously with this preprint).
%We are encouraged by this general problem being recognized by others in the community, and we emphasize that these works offer complementary neuroscientific contributions (different theoretical models of focus) and use different technical methodologies (ours is built on our prior work \cite{loaiza2017maximum}, theirs similarly \cite{LueckmannGoncalves_17}).
%These distinct methodologies and scientific investigations emphasize the increased importance and timeliness of both works.

 \subsubsection{Related work}\label{methods_related}

TODO: rewrite this whole section.

A closely related methodology, variational inference, uses optimization to approximate posterior distributions \cite{blei2017variational}.
Standard methods like stochastic gradient variational Bayes \cite{kingma2013auto} or black box variational inference \cite{ranganath2014black} simply do not work for inference in theoretical models of neural circuits, since they require tractable likelihoods $p(\mathbf{x} \mid \mathbf{z})$.
Work on likelihood-free variational inference (LFVI) \cite{tran2017hierarchical}, which like EPI seeks to do inference in models with intractable likelihoods, employs an addtional deep neural network as a ratio estimator, enabling an estimation of the optimization objective for variational inference.
Like LFVI, EPI can be framed as variational inference (see Section \ref{methods_VI}).
But, unlike LFVI, EPI uses a single deep network to learn a distribution and is optimized to produce an emergent property, rather than condition on data points.  
Optimizing the EPI objective is a technological challenge, the details of which we elaborate in Section \ref{methods_AL_opt}.  
Before going through those details, we ground this optimization in a toy example.
 We note that, during our preparation and early presentation of this work \cite{bittner2019degenerate, bittner2019examining}, another work has arisen with broadly similar goals: bringing statistical inference to mechanistic models of neural circuits (\cite{nonnenmacher2018sbi, desitler2019statistical, gonccalves2019training}, preprint posted simultaneously with this preprint).
We are encouraged by this general problem being recognized by others in the community, and we emphasize that these works offer complementary neuroscientific contributions (different theoretical models of focus) and use different technical methodologies (ours is built on our prior work \cite{loaiza2017maximum}, theirs similarly \cite{LueckmannGoncalves_17}).
These distinct methodologies and scientific investigations emphasize the increased importance and timeliness of both works.
